---
title: "Statistical Modelling Portfolio"
subtitle: "Motor Trend Car Road Tests — mtcars Dataset"
author: "Statistical Modeling Portfolio"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    template: custom_template.html
    latex_engine: xelatex
    number_sections: true
    code_folding: show
    fig_width: 9
    fig_height: 5
    fig_caption: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  comment   = "#>",
  fig.align = "center",
  out.width = "90%",
  dev       = "png",
  dev.args  = list(bg = "#ffffff")
)
```

```{r theme-setup, include=FALSE}
library(ggplot2)

theme_portfolio <- function(base_size = 14) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.background   = element_rect(fill = "#ffffff", colour = NA),
      panel.background  = element_rect(fill = "#f8f9fa", colour = NA),
      panel.border      = element_rect(fill = NA, colour = "#dee2e6", linewidth = 0.5),
      panel.grid.major  = element_line(colour = "#e9ecef", linewidth = 0.4),
      panel.grid.minor  = element_line(colour = "#f1f3f5", linewidth = 0.2),
      text              = element_text(colour = "#374151"),
      plot.title        = element_text(face = "bold", size = rel(1.15),
                                       colour = "#111827", hjust = 0,
                                       margin = margin(b = 5)),
      plot.subtitle     = element_text(size = rel(0.88), colour = "#6c757d",
                                       hjust = 0, margin = margin(b = 10)),
      plot.caption      = element_text(size = rel(0.76), colour = "#6c757d",
                                       hjust = 1),
      axis.title        = element_text(size = rel(0.88), colour = "#6c757d"),
      axis.text         = element_text(size = rel(0.84), colour = "#6c757d"),
      axis.ticks        = element_line(colour = "#dee2e6"),
      legend.background = element_rect(fill = "#ffffff", colour = "#dee2e6"),
      legend.key        = element_rect(fill = "#f8f9fa", colour = NA),
      legend.title      = element_text(size = rel(0.84), colour = "#6c757d"),
      legend.text       = element_text(size = rel(0.84), colour = "#374151"),
      legend.position   = "top",
      strip.background  = element_rect(fill = "#f1f3f5", colour = "#dee2e6"),
      strip.text        = element_text(face = "bold", size = rel(0.88),
                                       colour = "#374151"),
      plot.margin       = margin(14, 14, 14, 14)
    )
}

theme_set(theme_portfolio())

col_blue   <- "#2563eb"
col_green  <- "#059669"
col_red    <- "#dc2626"
col_amber  <- "#d97706"
col_violet <- "#7c3aed"
col_sky    <- "#0891b2"

two_cols  <- c(col_blue, col_red)
four_cols <- c(col_blue, col_red, col_green, col_amber)
```

---

> **Portfolio Objectives**
>
> | # | Objective | Methods Applied |
> |---|:----------|:----------------|
> | 1 | Describe probability as a foundation of statistical modelling, including inference and MLE | Log-likelihood visualisation; Normal, Bernoulli, and Multinomial distributions; AIC, BIC, and LRT derived from MLE |
> | 2 | Apply the appropriate GLM for a specific data context | OLS/Normal for continuous `mpg`; Binomial logit for binary `am`; Multinomial logit for categorical `cyl`; LDA as generative alternative |
> | 3 | Demonstrate model selection given a set of candidate models | AIC/BIC candidate comparison; LRT; stepwise selection; LOOCV; 10-fold CV; LASSO lambda tuning |
> | 4 | Express the results of statistical models to a general audience | Plain-language explanations throughout; odds ratio forest plot; predicted probability curves; non-technical conclusion |
> | 5 | Use programming software to fit and assess statistical models | R with `lm`, `glm`, `glmnet`, `nnet::multinom`, `MASS::lda`, `caret`, `pROC`, `car`, `lmtest`, `broom` |

---

# Dataset & Research Questions {#data}

## The mtcars Dataset

The mtcars dataset is a classic automotive dataset derived from the 1974 Motor Trend US magazine. 
It contains 32 car models and 11 performance and design variables, making it a popular benchmark
for statistical modeling, regression analysis, and exploratory data analysis.

Each row corresponds to a unique car model, and each column captures a specific mechanical or 
performance attribute. The dataset is useful for demonstrating relationships between
engine characteristics, fuel efficiency, and vehicle design choices. In my modelling workflow I would
like to answer these distinct research questions:

|   | Research Question | Outcome | Model(s) |
|---|:------------------|:--------|:---------|
| 1 | What predicts fuel efficiency? | `mpg` — continuous | Linear Regression, LASSO, Ridge |
| 2 | Can we classify transmission type? | `am` — binary (0/1) | Logistic Regression GLM, LDA |
| 3 | What predicts number of cylinders? | `cyl` — 3 categories | Multinomial Logistic Regression |

## Variable Dictionary

| Variable | Type | Description |
|:---------|:-----|:------------|
| `mpg`    | Continuous | Miles per gallon — **outcome 1** |
| `am`     | Binary (0/1) | Transmission: 0 = automatic, 1 = manual — **outcome 2** |
| `cyl`    | Categorical (4/6/8) | Number of cylinders — **outcome 3** |
| `disp`   | Continuous | Engine displacement (cu. in.) |
| `hp`     | Continuous | Gross horsepower |
| `drat`   | Continuous | Rear axle ratio |
| `wt`     | Continuous | Weight (1000 lbs) |
| `qsec`   | Continuous | Quarter mile time |
| `vs`     | Binary | Engine shape: 0 = V, 1 = straight |
| `gear`   | Count | Number of forward gears |
| `carb`   | Count | Number of carburettors |

## Loading Packages & Data

```{r load-packages}
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(broom) 
library(MASS)  
library(glmnet) 
library(nnet)   
library(kableExtra)  
library(scales)  
library(GGally) 
library(car)   
library(lmtest)  
library(pROC)   
library(ResourceSelection)
library(caret)
library(dplyr)
library(knitr)
library(gtsummary)
library(purrr)
library(tibble)
```

```{r load-data}
data(mtcars)

cars <- mtcars |>
  mutate(
    am  = factor(am,  levels = c(0, 1), labels = c("Automatic", "Manual")),
    cyl = factor(cyl, levels = c(4, 6, 8), labels = c("4-cyl", "6-cyl", "8-cyl")),
    vs  = factor(vs,  levels = c(0, 1), labels = c("V-shaped", "Straight")),
    gear = factor(gear),
    carb = factor(carb)
  )

cars$am_num <- as.integer(cars$am) - 1

cars |>
  summarise(
    Variable = names(cars),
    DataType = sapply(cars, class),
    Example  = sapply(cars, function(x) paste(head(x, 5), collapse = ", "))
  ) |>
  kable()
```

```{r data-summary}
mtcars |>
  select(mpg, disp, hp, drat, wt, qsec) |>
  map_df(~ as.numeric(summary(.x)), .id = "Statistic") |>
  mutate(Statistic = c("Min", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max")) |>
  select(Statistic, everything()) |>
  kable()

```

---

# Exploratory Data Analysis {#eda}

> **Objectives 1:** Examining distributions, relationships, and potential
> issues before modelling.

## Outcome Distributions

```{r eda-mpg, fig.cap="Fuel efficiency (mpg) is right-skewed with a mean of 20.1 mpg. Manual cars tend to achieve higher mpg."}
p1 <- ggplot(cars, aes(x = mpg, fill = am)) +
  geom_histogram(bins = 12, alpha = 0.8, colour = "white",
                 position = "dodge") +
  scale_fill_manual(values = two_cols) +
  labs(title = "Distribution of Fuel Efficiency (mpg)",
       subtitle = "Coloured by transmission type",
       x = "Miles per Gallon", y = "Count", fill = NULL) +
  theme_portfolio()
p1
```

```{r eda-cyl, fig.cap="Most cars in the dataset have 8 cylinders. Manual transmission is much more common in 4-cylinder cars."}
p2 <- ggplot(cars, aes(x = cyl, fill = am)) +
  geom_bar(position = "dodge", colour = "white", width = 0.65) +
  scale_fill_manual(values = two_cols) +
  labs(title = "Cylinder Count by Transmission Type",
       x = "Number of Cylinders", y = "Count", fill = NULL) +
  theme_portfolio()
p2
```

> **Explanation:**
> Automatic cars mostly appear in the lower mpg range (about 10--20 mpg),
> meaning they tend to use more fuel. Manual cars mostly appear in the higher
> mpg range (about 20--35 mpg), meaning they tend to be more fuel-efficient.

## Continuous Predictors

```{r eda-pairs, fig.cap="Pairs plot of key continuous variables. mpg shows strong negative correlations with wt, disp, and hp.", fig.height=7}
cars |>
  dplyr::select(mpg, wt, hp, disp, drat, qsec) |>
  ggpairs(
    upper = list(continuous = wrap("cor", size = 3.2, colour = "#374151")),
    diag  = list(continuous = wrap("densityDiag", fill = col_blue, alpha = 0.5)),
    lower = list(continuous = wrap("points", size = 1.5,
                                   colour = col_blue, alpha = 0.6))
  ) +
  labs(title = "Pairs Plot — Continuous Predictors") +
  theme_portfolio(base_size = 12)
```

### Correlation Analysis of Continuous Predictors

> The pairs plot summarizes the relationships among six continuous variables
> (*mpg*, *wt*, *hp*, *disp*, *drat*, and *qsec*) and reveals several strong,
> interpretable patterns. Overall, the predictors form two clear clusters:
> (1) engine size and power variables, and (2) efficiency and performance variables.
>
> **Relationships with mpg:**
> Fuel efficiency shows strong negative correlations with weight (r = -0.868),
> displacement (r = -0.848), and horsepower (r = -0.776). In contrast, mpg is
> positively correlated with rear axle ratio (r = 0.681) and quarter-mile time
> (r = 0.419).
>
> **Engine Size and Power Cluster:**
> The variables wt, disp, and hp are strongly positively correlated with one
> another (e.g., wt-disp: r = 0.888; hp-disp: r = 0.791). These strong
> intercorrelations reflect a clear "size/power" dimension and imply potential
> multicollinearity when included together in a regression model.

```{r eda-scatter, fig.cap="Weight is the single strongest predictor of mpg: heavier cars consume more fuel."}
cars %>%
  ggplot(aes(x = wt, y = mpg, colour = cyl, shape = am)) +
  geom_point(size = 3.5, alpha = 0.85) +
  geom_smooth(aes(group = 1), method = "lm", se = TRUE,
              colour = col_blue, fill = col_blue, alpha = 0.12, linewidth = 0.9) +
  scale_colour_manual(values = c(col_green, col_amber, col_red)) +
  labs(title    = "mpg vs Weight",
       subtitle = "Shape = transmission | Colour = cylinders",
       x = "Weight (1000 lbs)", y = "Miles per Gallon",
       colour = "Cylinders", shape = "Transmission") +
  theme_portfolio()
```

## Correlation Heatmap

```{r eda-corr, fig.cap="wt, disp, and hp are strongly negatively correlated with mpg. Multicollinearity between predictors is notable."}
cor_mat <- cor(mtcars %>% dplyr::select(mpg, disp, hp, drat, wt, qsec))

cor_mat %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "cor") %>%
  ggplot(aes(x = var2, y = var1, fill = cor)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = round(cor, 2)), size = 3.2, colour = "white", fontface = "bold") +
  scale_fill_gradient2(low = col_red, mid = "white", high = col_blue,
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation Matrix", x = NULL, y = NULL, fill = "r") +
  theme_portfolio() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid  = element_blank(),
        panel.background = element_rect(fill = "white"))
```

---

# Probability as a Foundation {#probability}

> **Objective 1:** Describing probability as a foundation of statistical modelling,
> including inference and Maximum Likelihood Estimation.

## Probability Distributions in This Portfolio

Every statistical model begins by specifying a **probability distribution** that
describes how the outcome was generated. The choice of distribution must match
the nature of the outcome variable:

| Outcome | Distribution | Rationale |
|:--------|:-------------|:----------|
| `mpg` — continuous | Normal: `Y_i ~ N(mu_i, sigma^2)` | Symmetric, continuous — OLS is both MLE and BLUE |
| `am` — binary | Bernoulli: `Y_i ~ Bernoulli(p_i)` | Two mutually exclusive outcomes |
| `cyl` — 3 categories | Multinomial(`p_i`) | Unordered polytomous outcome |

## Maximum Likelihood Estimation (MLE)

Maximum Likelihood Estimation (MLE) is a general framework for fitting
statistical models. The idea is to choose the parameter values (written here as
"theta‑hat") that make the observed data most probable under the model.

The likelihood function measures how well a given set of parameters explains
the data. Taking the logarithm of this function gives the log‑likelihood, which
is easier to work with because products become sums. The MLE is the parameter
value that maximizes this log‑likelihood.

For **linear regression** under the Normal error assumption, the MLE coincides
with the familiar Ordinary Least Squares (OLS) solution. In this case, the
coefficient estimates have a closed‑form expression based on matrix algebra.

For **logistic** and **multinomial** regression, there is no closed‑form
solution. Instead, the estimates are found numerically using the iterative
reweighted least squares (IRLS) algorithm, as implemented in functions such as
`glm()` and `multinom()`.

For **logistic** and **multinomial** regression, the solution is found numerically
via the IRLS algorithm inside `glm()` and `multinom()`.


## From MLE to Model Comparison

Once a model has been fitted by maximum likelihood, the value of the
log‑likelihood at the estimated parameters becomes the basis for comparing
different models.

Two widely used criteria are AIC and BIC. Both penalise model complexity to
discourage overfitting. AIC applies a lighter penalty and tends to favour more
flexible models, while BIC applies a stronger penalty and tends to prefer more
parsimonious ones. Both are computed from the log‑likelihood and the number of
estimated parameters.

The **Likelihood‑Ratio Test (LRT)** provides a formal hypothesis test for
comparing two nested models. It measures how much the log‑likelihood improves
when moving from a smaller model to a larger one. The resulting test statistic
follows a chi‑square distribution, with degrees of freedom equal to the
difference in the number of parameters between the two models.


```{r mle-viz, fig.cap="Log-likelihood for the Normal model fitted to mpg. The MLE (peak) equals the sample mean — the value of mu that makes the data most probable."}
mu_grid <- seq(15, 26, length.out = 500)
ll_vals <- sapply(mu_grid, function(mu) {
  sum(dnorm(mtcars$mpg, mean = mu, sd = sd(mtcars$mpg), log = TRUE))
})
mle_mu <- mean(mtcars$mpg)

data.frame(mu = mu_grid, ll = ll_vals) |>
  ggplot(aes(x = mu, y = ll)) +
  geom_line(colour = col_blue, linewidth = 1.4) +
  geom_vline(xintercept = mle_mu, linetype = "dashed",
             colour = col_red, linewidth = 0.9) +
  annotate("point", x = mle_mu, y = max(ll_vals), colour = col_red, size = 4) +
  annotate("label", x = mle_mu + 0.5, y = max(ll_vals) - 3,
           label = paste0("MLE: mu-hat = ", round(mle_mu, 2), " mpg"),
           colour = col_red, size = 4, fill = "white", label.size = 0.3) +
  labs(title    = "Log-Likelihood for Normal Model (outcome = mpg)",
       subtitle = "The MLE is the peak — the sample mean maximises the log-likelihood",
       x = "mu (mean mpg)", y = "Log-Likelihood") +
  theme_portfolio()
```

---

# Linear Regression {#linear}

> **Objectives 2, 3, 4, 5:** Predicting `mpg` (continuous outcome) using linear regression.
> Model selection via AIC/BIC and a stepwise search.

## Why Linear Regression?

`mpg` is a continuous, approximately symmetric outcome. Under the Normal
distribution assumption, OLS is the MLE and produces the Best Linear Unbiased
Estimator (BLUE).

Y_i = b_0 + b_1 x_{i1} + ... + b_p x_{ip} + e_i,  with  e_i ~ Normal(0, s^2)

## Fitting Candidate Models

```{r lm-candidates}
#specifying model
lm_spec <- 
  linear_reg() |>
  set_engine("lm")

#Fitting candidate models
lm_null <- lm_spec |> fit(mpg ~ 1, data = mtcars)
lm_eng  <- lm_spec |> fit(mpg ~ hp + disp, data = mtcars)
lm_core <- lm_spec |> fit(mpg ~ wt + hp + am, data = mtcars)
lm_full <- lm_spec |> fit(mpg ~ wt + hp + disp + drat +
                                qsec + vs + am + gear + carb,
                          data = mtcars)

# Bundling models
candidates <- list(
  Null   = lm_null,
  Engine = lm_eng,
  Core   = lm_core,
  Full   = lm_full
)

# Computing metrics for each model
model_summary <- map_dfr(
  candidates,
  function(m) {
    fit_obj <- extract_fit_engine(m)   # underlying lm object
    
    tibble(
      Predictors = length(coef(fit_obj)) - 1,
      R2         = glance(fit_obj)$r.squared,
      Adj_R2     = glance(fit_obj)$adj.r.squared,
      AIC        = AIC(fit_obj),
      BIC        = BIC(fit_obj),
      RMSE       = sqrt(mean(residuals(fit_obj)^2))
    )
  },
  .id = "Model"
)

model_summary |>
  mutate(
    R2     = round(R2, 3),
    Adj_R2 = round(Adj_R2, 3),
    AIC    = round(AIC, 2),
    BIC    = round(BIC, 2),
    RMSE   = round(RMSE, 3)
  ) |>
  kbl(
    caption = "Linear Regression — Candidate Model Comparison (disp removed after VIF check)",
    row.names = FALSE
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  row_spec(
    which.min(model_summary$AIC),
    bold = TRUE,
    background = "#f0fdf4",
    color = "#166534"
  )

```

>**Explanation:**The table above compares the four linear regression models after removing disp due to multicollinearity. The Core model (wt + hp + am) provides the best balance of performance and simplicity: it has strong R² and adjusted R², the lowest AIC, and competitive RMSE. Although the Full model fits slightly better, it uses many more predictors with diminishing returns, making the Core model the most persimonious model.


## Stepwise Selection

```{r lm-stepwise}
# Initial stepwise on full model — used for exploratory purposes only.
# The final lm_step is defined after multicollinearity is resolved below.
lm_full <- lm(
  mpg ~ wt + hp + disp + drat + qsec + vs + am + gear + carb,
  data = mtcars
)
lm_step_initial <- stepAIC(lm_full, direction = "both", trace = FALSE)
print(formula(lm_step_initial))
```

## Model Diagnostics

```{r lm-diagnostics, fig.cap="OLS diagnostic plots on the initial full model. These guide the multicollinearity diagnosis below.", fig.height=7}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1),
    bg = "#ffffff", col.axis = "#6c757d", col.lab = "#374151",
    col.main = "#111827")
plot(lm_full, which = 1:4, col = col_blue, pch = 16, cex = 0.8,
     cex.caption = 0.85)
par(mfrow = c(1, 1))
```

> **Regression Diagnostic Plots**
>
> The diagnostic plots provide a comprehensive assessment of the linear regression
> model assumptions, including linearity, normality of residuals, homoscedasticity,
> and the presence of influential observations.
>
> **Residuals vs Fitted:** No strong curvature, suggesting the linearity assumption
> is reasonably satisfied. A few observations (e.g., *Chrysler Imperial*, *Fiat 128*,
> *Toyota Corolla*) stand out with larger residuals, but the overall pattern does not
> indicate major model misspecification.
>
> **Normal Q-Q Plot:** The standardised residuals follow the theoretical Normal
> distribution fairly well with only mild deviations at the tails, indicating the
> normality assumption is acceptable.
>
> **Scale-Location:** The smoothing line is relatively horizontal, indicating
> approximately constant variance. No strong evidence of heteroscedasticity.
>
> **Cook's Distance:** A small number of cases (*Merc 230*, *Chrysler Imperial*,
> *Ford Pantera L*) show elevated Cook's distance values. None exceed the
> conventional cutoff of 1, but these points warrant closer inspection.

## Diagnosing Multicollinearity

Multicollinearity occurs when predictors are highly correlated, inflating
coefficient variance, widening confidence intervals, and making individual
effects unreliable. The **Variance Inflation Factor (VIF)** quantifies this:

VIF<sub>j</sub> = 1 / (1 − R<sub>j</sub><sup>2</sup>)


<p>
  The variance inflation factor (VIF) for predictor j is
  VIF<sub>j</sub> = 1 / (1 − R<sub>j</sub><sup>2</sup>),
  where R<sub>j</sub><sup>2</sup> is from regressing predictor j on all other predictors.
</p>

<p>
  A common rule of thumb is:
</p>
<ul>
  <li>VIF &lt; 5: acceptable</li>
  <li>5–10: warrants investigation</li>
  <li>&gt; 10: severe multicollinearity</li>
</ul>

```{r lm-corr-pred, fig.cap="Predictor correlation matrix. disp, wt, and hp all share correlations above 0.8 — the root cause of multicollinearity in the full model."}
pred_vars <- c("wt", "hp", "disp", "drat", "qsec")

cor(mtcars[, pred_vars]) |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  ggplot(aes(x = var2, y = var1, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = round(r, 2),
                colour = abs(r) > 0.65),
            size = 4.2, fontface = "bold") +
  scale_fill_gradient2(low = col_red, mid = "white", high = col_blue,
                       midpoint = 0, limits = c(-1, 1)) +
  scale_colour_manual(values = c("FALSE" = "#374151", "TRUE" = "#ffffff"),
                      guide = "none") +
  scale_x_discrete(position = "top") +
  labs(title    = "Predictor Correlation Matrix Root Cause of Multicollinearity",
       subtitle = "Pairs with |r| > 0.8 carry redundant information",
       x = NULL, y = NULL, fill = "r") +
  theme_portfolio() +
  theme(panel.grid       = element_blank(),
        panel.background = element_rect(fill = "white"))
```

```{r lm-vif-full}
vif_full <- car::vif(lm_full) |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  setNames(c("Term", "VIF")) |>
  mutate(VIF    = round(VIF, 3),
         Status = case_when(VIF > 10 ~ "Severe",
                            VIF > 5  ~ "Moderate",
                            TRUE     ~ "OK"))

vif_full |>
  kbl(caption = "Step 1 — VIF in the Full Model: Severe multicollinearity detected",
      row.names = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  row_spec(which(vif_full$VIF > 10), background = "#fff1f2", color = "#9f1239") |>
  row_spec(which(vif_full$VIF > 5 & vif_full$VIF <= 10),
           background = "#fffbeb", color = "#92400e")
```

## Solving Multicollinearity

Three complementary strategies are applied to resolve the collinearity between
`disp`, `wt`, and `hp`:

**Dropping the most redundant predictor.** `disp` has the highest VIF and is
the most strongly correlated with both `wt` and `hp`. It is removed before
stepwise selection.

**Stepwise selection.** With `disp` excluded, `stepAIC()` selects among the
remaining predictors using AIC, further pruning any that do not contribute unique
information.

**Ridge regression.** Ridge shrinks collinear coefficients toward each other
rather than to zero, directly mitigating the variance inflation caused by
correlated predictors. Applied in the regularisation section.

```{r lm-reduced}
library(MASS)

#Fitting the reduced model
lm_reduced <- lm(mpg ~ wt + hp + drat + qsec + vs + am + gear + carb,
                 data = mtcars)

# Running stepwise AIC
lm_step <- stepAIC(lm_reduced, direction = "both", trace = FALSE)

cat("Formula after stepwise selection:\n")
print(formula(lm_step))


```

```{r lm-vif-solved}
vif_solved <- car::vif(lm_step) |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  setNames(c("Term", "VIF")) |>
  mutate(VIF    = round(VIF, 3),
         Status = case_when(VIF > 10 ~ "Severe",
                            VIF > 5  ~ "Moderate",
                            TRUE     ~ "OK"))

vif_solved |>
  kbl(caption = "Step 2 — VIF After Removing disp and Re-running Stepwise: All clear",
      row.names = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  row_spec(which(vif_solved$Status == "OK"),
           background = "#f0fdf4", color = "#166534")
```

```{r lm-vif-compare-plot, fig.cap="VIF before and after multicollinearity resolution. All predictors in the final model fall comfortably below VIF = 5."}
bind_rows(
  vif_full   |> mutate(Model = "Full Model (with disp)"),
  vif_solved |> mutate(Model = "Final Model (disp removed + stepwise)")
) |>
  mutate(Term = fct_reorder(Term, VIF, .fun = max)) |>
  ggplot(aes(x = VIF, y = Term, fill = Model)) +
  geom_col(position = "dodge", width = 0.65) +
  geom_vline(xintercept = 5,  linetype = "dashed",
             colour = col_red,   linewidth = 0.9) +
  geom_vline(xintercept = 10, linetype = "dotted",
             colour = col_amber, linewidth = 0.9) +
  scale_fill_manual(values = c(
    "Full Model (with disp)"                = col_blue,
    "Final Model (disp removed + stepwise)" = col_green)) +
  annotate("text", x = 5.3,  y = 0.5, label = "VIF = 5",
           size = 3, colour = col_red,   hjust = 0, fontface = "italic") +
  annotate("text", x = 10.3, y = 0.5, label = "VIF = 10",
           size = 3, colour = col_amber, hjust = 0, fontface = "italic") +
  labs(title    = "Multicollinearity Resolution: VIF Before and After",
       subtitle = "Removing disp and re-running stepwise brings all VIF values below 5",
       x = "VIF", y = NULL, fill = NULL) +
  theme_portfolio()
```

> **Explanation:** The full model had severe multicollinearity because `disp`,
> `wt`, and `hp` all measure engine size and power, they tell the model the same
> thing three different ways. Removing `disp` and re-running stepwise selection
> brings all VIF values below 5, meaning each predictor now makes an independent
> contribution.

## Final Model Summary

```{r lm-summary}
summary(lm_step)
```

## Coefficient Table

```{r lm-coef-table}
tidy(lm_step, conf.int = TRUE) |>
  mutate(across(c(estimate, std.error, conf.low, conf.high), \(x) round(x, 3)),
         p.value = round(p.value, 4),
         Sig = case_when(
           p.value < 0.001 ~ "***", p.value < 0.01 ~ "**",
           p.value < 0.05  ~ "*",   p.value < 0.10 ~ ".",
           TRUE ~ ""
         )) |>
  rename(Term = term, Estimate = estimate, SE = std.error,
         `t value` = statistic, `p-value` = p.value,
         `2.5%` = conf.low, `97.5%` = conf.high) |>
  kbl(caption = "Final Linear Regression Model — Coefficients (disp removed, VIF resolved)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

> **Explanation:** According to the above linear regression model:
>
> `wt`: An increase in weight by 1,000 leads to a decrease in miles by gallon by 3.9 miles keeping all other variables constant.
>
> `qsec`: An increase in quarter mile time by 1 unit leads to an increase in miles by gallon by 1.2 miles, keeping all other variables constant.
>
> `am`: A manual car achieves 2.936miles more than an automatic car.
>`Intercept`: For an automatic car keeping all other variables constant it achieves an average mpg of 9.618

## Prediction interval
```{r}
#sample data point
new_point <- tibble(
  wt   = 3.0,
  qsec = 17.0,
  am   = 1
)

predict(lm_step, newdata = new_point, interval = "prediction")



```
>**Explanation:** We are 95% confident that for a manual car with a weight of 3,000 pounds, quarter mile time of 17.0, the predicted miles per gallon is between 16.34 miles and 26.9 miles.

```{r lm-resid-final, fig.cap="Residual distribution for the final model (disp removed). Residuals remain approximately Normal confirming OLS assumptions still hold."}
data.frame(resid = resid(lm_step)) |>
  ggplot(aes(x = resid)) +
  geom_histogram(aes(y = after_stat(density)), bins = 15,
                 fill = col_green, alpha = 0.6, colour = "white") +
  geom_density(linewidth = 1, colour = "#111827") +
  stat_function(fun  = dnorm,
                args = list(mean = mean(resid(lm_step)),
                            sd   = sd(resid(lm_step))),
                colour = col_red, linetype = "dashed", linewidth = 1.1) +
  labs(title    = "Residual Distribution — Final Model (Multicollinearity Resolved)",
       subtitle = "Black = empirical density | Red dashed = fitted Normal",
       x = "Residual", y = "Density") +
  theme_portfolio()
```

```{r lm-bp}
bptest(lm_step)
```

```{r lm-actual-fitted, fig.cap="Actual vs. fitted mpg. Points close to the dashed line indicate good model fit."}
data.frame(actual = mtcars$mpg, fitted = fitted(lm_step),
           am = cars$am) |>
  ggplot(aes(x = fitted, y = actual, colour = am)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "#6c757d", linewidth = 0.8) +
  geom_point(size = 3.5, alpha = 0.85) +
  scale_colour_manual(values = two_cols) +
  labs(title  = "Actual vs. Fitted — mpg (Linear Regression)",
       x = "Fitted mpg", y = "Actual mpg", colour = "Transmission") +
  theme_portfolio()
```

> **Explanation:** The stepwise model explains
> **`r round(summary(lm_step)$adj.r.squared * 100, 1)`%** of the variation in
> fuel efficiency. Weight (`wt`) is the strongest predictor, heavier cars
> consume more fuel. Horsepower (`hp`) also significantly reduces mpg, while
> quarter-mile time (`qsec`) and transmission type (`am`) each contribute
> meaningfully after accounting for the others.

---

# LASSO and Ridge Regression {#regularisation}

> **Objectives 2, 3, 5:**

Regularised regression adds a penalty on the size of the coefficient vector to
reduce overfitting and improve prediction.

**Ridge regression (L2 penalty)** shrinks all coefficients toward zero but never
sets any of them exactly to zero. It is most useful when many predictors have
small, shared influence and multicollinearity is present.

**LASSO regression (L1 penalty)** can shrink some coefficients all the way to
zero, effectively performing variable selection. It is helpful when only a
subset of predictors is truly important.

In both methods, the strength of the penalty is controlled by a tuning
parameter (lambda), which is typically chosen by cross‑validation to balance
model complexity and predictive accuracy.

```{r reg-prep}
X_reg <- model.matrix(mpg ~ wt + hp + disp + drat + qsec + vs + am + gear + carb,
                      data = mtcars)[, -1]
y_reg <- mtcars$mpg
cat("Design matrix:", nrow(X_reg), "rows x", ncol(X_reg), "columns\n")
```

## LASSO

```{r lasso-cv, fig.cap="LASSO: 10-fold CV MSE across log(lambda). Left dashed line = lambda.min; right = lambda.1se (more regularised)."}
set.seed(42)
cv_lasso <- cv.glmnet(X_reg, y_reg, alpha = 1, nfolds = 10)
plot(cv_lasso)
title("LASSO — 10-Fold CV MSE vs log(lambda)", line = 2.5)
cat("Best lambda (min):", round(cv_lasso$lambda.min, 4),
    "\nBest lambda (1se):", round(cv_lasso$lambda.1se, 4))
```

```{r lasso-coefs}
coef(cv_lasso, s = "lambda.min") |>
  as.matrix() |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  setNames(c("Term", "Coefficient")) |>
  filter(Term != "(Intercept)") |>
  mutate(Coefficient = round(Coefficient, 4),
         Status = ifelse(Coefficient == 0, "Zeroed (LASSO)", "Retained")) |>
  arrange(desc(abs(Coefficient))) |>
  kbl(caption = paste0("LASSO Coefficients (lambda = ",
                       round(cv_lasso$lambda.min, 4), ")"),
      row.names = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Ridge

```{r ridge-cv, fig.cap="Ridge: all coefficients are retained but shrunk toward zero as lambda increases."}
set.seed(42)
cv_ridge <- cv.glmnet(X_reg, y_reg, alpha = 0, nfolds = 10)
plot(cv_ridge)
title("Ridge — 10-Fold CV MSE vs log(lambda)", line = 2.5)
cat("Best lambda (min):", round(cv_ridge$lambda.min, 4))
```

## OLS vs LASSO vs Ridge

```{r reg-compare}
set.seed(42)
idx  <- sample(nrow(mtcars), 0.75 * nrow(mtcars))

X_tr <- X_reg[idx, ];   X_te <- X_reg[-idx, ]
y_tr <- y_reg[idx];     y_te <- y_reg[-idx]

ols_tr <- lm(mpg ~ wt + hp + disp + drat + qsec + vs + am + gear + carb,
             data = mtcars[idx, ])

rmse <- function(a, p) round(sqrt(mean((a - p)^2)), 3)

data.frame(
  Method      = c("OLS", "LASSO", "Ridge"),
  `Test RMSE` = c(
    rmse(y_te, predict(ols_tr, mtcars[-idx, ])),
    rmse(y_te, as.numeric(predict(cv.glmnet(X_tr, y_tr, alpha = 1),
                                  X_te, s = "lambda.min"))),
    rmse(y_te, as.numeric(predict(cv.glmnet(X_tr, y_tr, alpha = 0),
                                  X_te, s = "lambda.min")))
  ),
  Notes = c("No regularisation", "L1 — zeroes weak predictors",
             "L2 — shrinks all, keeps all")
) |>
  kbl(caption = "Test-Set RMSE: OLS vs LASSO vs Ridge (75/25 split)",
      row.names = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

> **Explanation:** LASSO automatically identified the most important predictors
> and set others to zero, confirming that `wt`, `hp`, and `qsec` carry most of
> the signal. Ridge retained all predictors with smaller coefficients. Both
> regularised models produce comparable or better out-of-sample accuracy than OLS.

---

# Logistic Regression GLM {#logistic}

> **Objectives 1, 2, 3, 4:** Predicting binary `am` (transmission type:
> automatic vs manual). The Binomial GLM with logit link is the appropriate model.

## Why Logistic Regression?

`am` is binary, so the Normal distribution is inappropriate. The Bernoulli
distribution is the natural choice, with the **logit** as the canonical link:

| Component   | Logistic Regression |
|:----------- |:--------------------|
| **Random**      | Y_i follows a Bernoulli(p_i) distribution |
| **Systematic**  | eta_i = x_i^T b |
| **Link**        | logit(p_i) = log(p_i / (1 - p_i)) = eta_i |

## A Note on Complete Separation

With only n = 32 observations, adding too many predictors to a logistic model
causes complete separation, the model perfectly predicts every training
observation, driving the log-likelihood to -infinity, coefficient estimates to
plus or minus infinity, and producing deviance approximately 0 with meaningless AIC/BIC values.
This is not a sign of a good model; it is a sign of overfitting.

The rule of thumb for logistic regression is at least 10--20 events per
predictor (EPP). With 19 manual cars (the minority class), we can support at
most 1--2 predictors reliably. The candidate models are therefore kept
deliberately parsimonious.

## Candidate Models & Selection

```{r log-candidates}
log_null  <- glm(am ~ 1,       data = cars, family = binomial())
log_wt    <- glm(am ~ wt,      data = cars, family = binomial())
log_hp    <- glm(am ~ hp,      data = cars, family = binomial())
log_core  <- glm(am ~ wt + hp, data = cars, family = binomial())

log_overfit <- glm(am ~ wt + hp + qsec, data = cars, family = binomial())

cat("Deviance check \n")
cat("Core  (2 predictors):", round(deviance(log_core),    2), "\n")
cat("Overfit (3 predictors):", round(deviance(log_overfit), 2),
    "<-- near zero = complete separation\n")

map_dfr(list(Null = log_null, `wt only` = log_wt,
             `hp only` = log_hp, `wt + hp (Core)` = log_core),
        \(m) data.frame(
          Predictors = length(coef(m)) - 1,
          Deviance   = round(deviance(m), 2),
          AIC        = round(AIC(m), 2),
          BIC        = round(BIC(m), 2),
          Note       = ifelse(deviance(m) < 1, "Complete separation", "OK")
        ), .id = "Model") |>
  kbl(caption = paste0("Logistic Regression — Candidate Model Comparison",
                       " (n=32, EPP limits to 1-2 predictors)"),
      row.names = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  row_spec(4, bold = TRUE, background = "#f0fdf4", color = "#166534")
```

```{r log-lrt}
cat("=== LRT: Null vs wt only ===\n")
print(anova(log_null, log_wt, test = "Chisq"))
cat("\n=== LRT: wt only vs wt + hp (Core) ===\n")
print(anova(log_wt, log_core, test = "Chisq"))
cat("\nConclusion: wt + hp significantly improves on wt alone.",
    "\nAdding a 3rd predictor causes complete separation — Core model selected.\n")
```

```{r log-separation-plot, fig.cap="Weight alone almost completely separates automatic from manual cars. This near-perfect predictability is why models with 3+ predictors achieve deviance close to zero."}
ggplot(cars, aes(x = wt, y = as.numeric(am) - 1, colour = am)) +
  geom_point(size = 4, alpha = 0.85) +
  geom_smooth(method = "glm", method.args = list(family = binomial()),
              aes(group = 1), colour = "#111827",
              se = TRUE, fill = "#dee2e6", alpha = 0.2, linewidth = 1) +
  scale_colour_manual(values = two_cols) +
  scale_y_continuous(breaks = c(0, 1),
                     labels = c("0 — Automatic", "1 — Manual")) +
  labs(title    = "Near-Complete Separation: Weight vs Transmission",
       subtitle = "Heavier cars are almost exclusively automatic — near-perfect predictability",
       x = "Weight (1000 lbs)", y = "Transmission (am)",
       colour = NULL) +
  theme_portfolio()
```

```{r log-final}
log_final <- log_core
summary(log_final)
```

## Odds Ratios

```{r log-or}
tidy(log_final, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(across(c(estimate, conf.low, conf.high), \(x) round(x, 3)),
         p.value = round(p.value, 4),
         Sig = case_when(p.value < 0.001 ~ "***", p.value < 0.01 ~ "**",
                         p.value < 0.05  ~ "*",   p.value < 0.10 ~ ".",
                         TRUE ~ "")) |>
  rename(Predictor=term, OR=estimate, `2.5%`=conf.low,
         `97.5%`=conf.high, `p-value`=p.value) |>
  kbl(caption = "Logistic Regression — Odds Ratios with 95% CIs") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r log-or-plot, fig.cap="Forest plot of odds ratios. OR < 1 means the predictor is associated with lower odds of a manual transmission."}
tidy(log_final, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(term = fct_reorder(term, estimate),
         dir  = ifelse(estimate >= 1, "Higher odds (Manual)", "Lower odds (Manual)")) |>
  ggplot(aes(x = estimate, y = term, colour = dir)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "#6c757d") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2, linewidth = 0.8) +
  geom_point(size = 4) +
  scale_colour_manual(values = c(col_blue, col_red)) +
  scale_x_log10() +
  labs(title  = "Odds Ratio Forest Plot — Logistic Regression",
       x = "Odds Ratio (log scale)", y = NULL, colour = NULL) +
  theme_portfolio()
```

## Logistic Model Diagnostics

```{r log-hoslem}
hl <- hoslem.test(cars$am_num, fitted(log_final), g = 8)
print(hl)
cat("\np =", round(hl$p.value, 3),
    if (hl$p.value > 0.05) "— model fits well" else "— evidence of misfit")
```

```{r log-vif}
log_vif_aux <- lm(am_num ~ wt + hp, data = cars)

car::vif(log_vif_aux) |>
  as.data.frame() |>
  rownames_to_column("Predictor") |>
  setNames(c("Predictor", "VIF")) |>
  mutate(VIF    = round(VIF, 3),
         Status = case_when(VIF > 10 ~ "Severe",
                            VIF > 5  ~ "Moderate",
                            TRUE     ~ "OK")) |>
  kbl(caption = "VIF — Logistic Regression Predictors (wt and hp)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r log-corr-pred, fig.cap="Correlation between wt and hp. Moderate correlation exists but VIF confirms it stays within an acceptable range for the logistic model."}
data.frame(wt = mtcars$wt, hp = mtcars$hp) |>
  ggplot(aes(x = wt, y = hp)) +
  geom_point(size = 3, colour = col_blue, alpha = 0.75) +
  geom_smooth(method = "lm", colour = col_red, fill = col_red,
              alpha = 0.12, linewidth = 0.9, se = TRUE) +
  annotate("label", x = 4.5, y = 100, fill = "white", label.size = 0.3,
           label = paste0("r = ", round(cor(mtcars$wt, mtcars$hp), 3)),
           size = 4, colour = col_red, fontface = "bold") +
  labs(title    = "Collinearity Check: Weight vs Horsepower",
       subtitle = "Moderate correlation (r ~ 0.66) — acceptable for logistic regression",
       x = "Weight (1000 lbs)", y = "Horsepower") +
  theme_portfolio()
```

```{r log-roc, fig.cap="ROC curve for the logistic regression model. AUC close to 1 indicates excellent discrimination."}
roc_log <- roc(cars$am_num, fitted(log_final),
               levels = c(0, 1), direction = "<", quiet = TRUE)
auc_log <- as.numeric(auc(roc_log))

ggroc(roc_log, colour = col_blue, linewidth = 1.4) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed",
              colour = "#6c757d", linewidth = 0.8) +
  annotate("label", x = 0.35, y = 0.15, fill = "white", label.size = 0.3,
           label = paste0("AUC = ", round(auc_log, 3)),
           size = 4.5, colour = col_blue, fontface = "bold") +
  labs(title = "ROC Curve — Logistic Regression (Transmission Type)",
       x = "1 - Specificity (FPR)", y = "Sensitivity (TPR)") +
  theme_portfolio()
```

```{r log-confmat}
pred_class <- ifelse(fitted(log_final) > 0.5, "Manual", "Automatic")
conf_log   <- table(Predicted = pred_class, Actual = cars$am)
print(conf_log)
log_acc <- sum(diag(conf_log)) / sum(conf_log)
cat("\nAccuracy:", round(log_acc * 100, 1), "% | AUC:", round(auc_log, 3))
```

> **Explanation:** The model correctly classifies
> **`r round(log_acc * 100, 1)`%** of cars as automatic or manual. Weight is
> the strongest predictor — heavier cars are much more likely to have automatic
> transmission. Each additional 1,000 lbs reduces the odds of manual transmission
> by approximately
> **`r round((1 - tidy(log_final, exponentiate=TRUE)$estimate[2]) * 100, 0)`%**,
> holding horsepower constant. A three-predictor model was deliberately avoided:
> with only 19 manual cars, adding a third predictor causes complete separation
> and produces an unreliable, overfit model.

---

# Multinomial Logistic Regression {#multinomial}

> **Objectives 1, 2, 4, 5:** Predicting `cyl` (4, 6, or 8 cylinders — three
> unordered categories). Multinomial regression is required because cylinder
> count has no meaningful numeric spacing.

## Why Multinomial?

With three outcome categories, two log‑odds equations are estimated relative to a
reference category (here, "4‑cyl").

The model expresses the log of the probability of category k relative to the
probability of the reference category as a linear function of the predictors:

log( P(Y = k) / P(Y = "4-cyl") ) =
  b_{k0} + b_{k1} x_1 + ... + b_{kp} x_p,
  for k in { "6-cyl", "8-cyl" }.

This allows the model to compare each non‑reference category directly to the
baseline while using a single unified framework.

```{r multi-fit}
cars$cyl <- relevel(cars$cyl, ref = "4-cyl")

multi_model <- nnet::multinom(
  cyl ~ wt + hp + disp,
  data  = cars,
  trace = FALSE
)
summary(multi_model)
```

## P-Values and Z-Statistics

```{r multi-pvals}
z_mat <- summary(multi_model)$coefficients /
         summary(multi_model)$standard.errors
p_mat <- 2 * (1 - pnorm(abs(z_mat)))

coef_m <- summary(multi_model)$coefficients
se_m   <- summary(multi_model)$standard.errors

map_dfr(rownames(coef_m), function(cat) {
  data.frame(
    Category = cat,
    Term     = colnames(coef_m),
    Estimate = round(coef_m[cat, ], 4),
    SE       = round(se_m[cat, ],   4),
    z        = round(z_mat[cat, ],  3),
    p.value  = round(p_mat[cat, ],  4),
    row.names = NULL
  )
}) |>
  filter(Term != "(Intercept)") |>
  mutate(Sig = case_when(
    p.value < 0.001 ~ "***", p.value < 0.01 ~ "**",
    p.value < 0.05  ~ "*",   p.value < 0.10 ~ ".",
    TRUE ~ ""
  )) |>
  kbl(caption = "Multinomial Regression Coefficients (ref = 4-cyl)",
      row.names = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Predicted Probabilities

```{r multi-probs, fig.cap="Predicted probability of each cylinder category by weight. Heavier cars are much more likely to have 8 cylinders.", echo=FALSE}
wt_grid <- data.frame(
  wt   = seq(1.5, 5.5, length.out = 100),
  hp   = mean(mtcars$hp),
  disp = mean(mtcars$disp)
)

probs <- predict(multi_model, newdata = wt_grid, type = "probs")

bind_cols(wt_grid, as.data.frame(probs)) |>
  pivot_longer(cols = c(`4-cyl`, `6-cyl`, `8-cyl`),
               names_to = "cylinders", values_to = "probability") |>
  ggplot(aes(x = wt, y = probability, colour = cylinders)) +
  geom_line(linewidth = 1.4) +
  scale_colour_manual(values = c(col_green, col_amber, col_red)) +
  labs(title    = "Predicted Cylinder Category Probabilities vs Weight",
       subtitle = "hp and disp held at their means",
       x = "Weight (1000 lbs)", y = "Predicted Probability",
       colour = "Cylinders") +
  theme_portfolio()
```

```{r multi-confmat}
multi_pred <- predict(multi_model, cars)
conf_multi  <- table(Predicted = multi_pred, Actual = cars$cyl)
print(conf_multi)
multi_acc <- sum(diag(conf_multi)) / sum(conf_multi)
cat("\nMultinomial Accuracy:", round(multi_acc * 100, 1), "%\n")
```

> **Explanation:** The model correctly classifies
> **`r round(multi_acc * 100, 1)`%** of cars into the correct cylinder category.
> Weight, horsepower, and displacement all contribute to the classification.
> Heavier, more powerful cars with larger engines are overwhelmingly likely to
> have 8 cylinders.

---

# Linear Discriminant Analysis {#lda}

> **Objectives 2, 3, 4, 5:** LDA is a generative classifier that models each
> class using a multivariate Normal distribution with a shared covariance matrix,
> then classifies via Bayes' theorem.

## Probability Foundation
For each class k, the predictor vector x is assumed to follow a Normal
distribution with its own class‑specific mean (mu_k) and a covariance matrix
(Sigma) that is shared across all classes.

The class probability for Y = k given x is proportional to the prior probability
of class k (pi_k) multiplied by the class‑specific density f_k(x), and then
normalised by the sum of these values across all classes.

Because all classes share the same covariance matrix Sigma, the resulting
decision boundary is linear. This is the key property that distinguishes
Linear Discriminant Analysis from its quadratic counterpart.

```{r lda-fit}
lda_model <- lda(am ~ wt + hp + qsec, data = cars)
lda_model
```

```{r lda-scores, fig.cap="Distribution of LD1 scores by transmission type. Clear separation confirms strong discriminability."}
lda_pred   <- predict(lda_model, cars)
lda_scores <- data.frame(LD1 = lda_pred$x[, 1], am = cars$am)

ggplot(lda_scores, aes(x = LD1, fill = am)) +
  geom_density(alpha = 0.6, colour = NA) +
  geom_rug(aes(colour = am), sides = "b", alpha = 0.5) +
  scale_fill_manual(values  = two_cols) +
  scale_colour_manual(values = two_cols) +
  labs(title    = "LDA Discriminant Score (LD1) by Transmission Type",
       subtitle = "Greater separation = stronger discriminability",
       x = "Linear Discriminant Score (LD1)",
       y = "Density", fill = NULL, colour = NULL) +
  theme_portfolio()
```

## LDA Performance

```{r lda-performance}
conf_lda <- table(Predicted = lda_pred$class, Actual = cars$am)
print(conf_lda)
lda_acc <- sum(diag(conf_lda)) / sum(conf_lda)

roc_lda <- roc(cars$am_num,
               lda_pred$posterior[, "Manual"],
               levels = c(0, 1), direction = "<", quiet = TRUE)
auc_lda <- as.numeric(auc(roc_lda))
cat("\nLDA Accuracy:", round(lda_acc * 100, 1),
    "% | AUC:", round(auc_lda, 3))
```

```{r lda-roc-comp, fig.cap="ROC comparison: Logistic Regression and LDA achieve similar discrimination for transmission type classification."}
ggroc(list(`Logistic Regression` = roc_log, LDA = roc_lda), linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed",
              colour = "#6c757d", linewidth = 0.8) +
  scale_colour_manual(values = c(col_blue, col_red)) +
  annotate("text", x = 0.3, y = 0.18, hjust = 0, size = 3.8,
           label = paste0("Logistic AUC = ", round(auc_log, 3),
                          "\nLDA AUC = ", round(auc_lda, 3))) +
  labs(title  = "ROC Curves — Logistic Regression vs LDA",
       x = "1 - Specificity", y = "Sensitivity", colour = NULL) +
  theme_portfolio()
```

## Normality Assumption Check

```{r lda-qq, fig.cap="Q-Q plots for LDA predictors. Approximate normality is sufficient for LDA to perform well.", fig.height=4}
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1),
    bg = "#ffffff", col.axis = "#6c757d", col.lab = "#374151")
for (v in c("wt", "hp", "qsec")) {
  qqnorm(mtcars[[v]], main = paste("Q-Q:", v),
         pch = 16, col = col_blue, cex = 0.85)
  qqline(mtcars[[v]], col = col_red, lwd = 1.5)
}
par(mfrow = c(1, 1))
```

> **Explanation:** LDA achieved an AUC of **`r round(auc_lda, 3)`**, very
> close to logistic regression (**`r round(auc_log, 3)`**). Both approaches
> correctly classify the majority of cars by transmission type. Logistic
> regression is preferred when distributional assumptions are uncertain, while
> LDA may be favoured when normality holds.

---

# Cross-Validation & Model Selection {#cv}

> **Objective 3, 5:** Cross-validation provides honest out-of-sample
> performance estimates and is used to make the final model selection decision.

## Why Cross-Validation?

With only n = 32 observations, a single train/test split is unreliable.
**Leave-One-Out Cross-Validation (LOOCV)** avoids this by repeatedly training
the model on 31 cars and predicting the one left out:

```{r cv-setup}
set.seed(123)
ctrl_loocv <- trainControl(method = "LOOCV")
ctrl_cv10  <- trainControl(method = "cv", number = 10,
                            classProbs      = TRUE,
                            summaryFunction = twoClassSummary,
                            savePredictions = "final")
```

## LOOCV (Linear Models)

```{r cv-lm}
cv_ols <- train(mpg ~ wt + hp + qsec + am, data = cars,
                method    = "lm",
                trControl = ctrl_loocv)

cv_las <- train(mpg ~ wt + hp + disp + drat + qsec + vs + am + gear + carb,
                data     = cars,
                method   = "glmnet",
                tuneGrid = expand.grid(alpha = 1,
                                       lambda = 10^seq(-3, 1, length.out = 40)),
                trControl = ctrl_loocv)

cv_rid <- train(mpg ~ wt + hp + disp + drat + qsec + vs + am + gear + carb,
                data     = cars,
                method   = "glmnet",
                tuneGrid = expand.grid(alpha = 0,
                                       lambda = 10^seq(-2, 2, length.out = 40)),
                trControl = ctrl_loocv)

data.frame(
  Model       = c("OLS (stepwise)", "LASSO", "Ridge"),
  LOOCV_RMSE  = round(c(cv_ols$results$RMSE,
                          min(cv_las$results$RMSE),
                          min(cv_rid$results$RMSE)), 3),
  LOOCV_R2    = round(c(cv_ols$results$Rsquared,
                          cv_las$results$Rsquared[which.min(cv_las$results$RMSE)],
                          cv_rid$results$Rsquared[which.min(cv_rid$results$RMSE)]), 3)
) |>
  kbl(caption = "LOOCV Performance — Regression Models (outcome = mpg)",
      row.names = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  row_spec(1, bold = TRUE, background = "#f0fdf4", color = "#166534")
```

## 10-Fold CV — Classification Models

```{r cv-classif}
cars_cv <- cars |>
  mutate(am_cv = factor(am_num, levels = c(0, 1),
                         labels = c("Automatic", "Manual")))

set.seed(123)
cv_log_cls <- train(
  am_cv ~ wt + hp,
  data      = cars_cv,
  method    = "glm",
  family    = "binomial",
  trControl = ctrl_cv10,
  metric    = "ROC"
)

set.seed(123)
cv_lda_cls <- train(
  am_cv ~ wt + hp + qsec,
  data      = cars_cv,
  method    = "lda",
  trControl = ctrl_cv10,
  metric    = "ROC"
)

data.frame(
  Model    = c("Logistic Regression", "LDA"),
  CV_AUC   = round(c(cv_log_cls$results$ROC,
                      cv_lda_cls$results$ROC), 3),
  CV_Sens  = round(c(cv_log_cls$results$Sens,
                      cv_lda_cls$results$Sens), 3),
  CV_Spec  = round(c(cv_log_cls$results$Spec,
                      cv_lda_cls$results$Spec), 3)
) |>
  kbl(caption = "10-Fold CV Performance — Classification Models (outcome = am)",
      row.names = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  row_spec(1, bold = TRUE, background = "#f0fdf4", color = "#166534")
```

## Final Model Selection

```{r final-selection}
tribble(
  ~Outcome, ~`Selected Model`, ~`CV Metric`, ~Reason,
  "mpg (continuous)",   "OLS — stepwise",         paste0("LOOCV RMSE = ", round(cv_ols$results$RMSE, 2)),  "Lowest error, highest interpretability, all assumptions met",
  "am (binary)",        "Logistic Regression",    paste0("CV AUC = ", round(cv_log_cls$results$ROC, 3)),    "Highest AUC, directly interpretable odds ratios",
  "cyl (3 categories)", "Multinomial Regression", "Accuracy = 96.9%",                                       "Only model appropriate for unordered polytomous outcome"
) |>
  kbl(caption = "Final Model Selection Summary") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) |>
  column_spec(2, bold = TRUE, color = "#166534")
```

---

# Model Diagnostics Summary {#diagnostics}

> **Objectives 3 & 5:** Consolidated diagnostics across all fitted models.

| Model | Check | Tool | Result |
|:------|:------|:-----|:-------|
| Linear Regression | Normality of residuals | Q-Q plot, histogram | Approximately Normal |
| Linear Regression | Homoscedasticity | Breusch-Pagan test | No violation |
| Linear Regression | Multicollinearity | VIF + predictor removal | disp dropped (VIF > 10, r > 0.85 with wt); final model all VIF < 5 |
| Linear Regression | Influential observations | Cook's distance | No extreme points |
| LASSO / Ridge | Optimal lambda | 10-fold CV path | lambda selected automatically |
| Logistic Regression | Complete separation | EPP check + deviance | 3+ predictors cause separation; core 2-predictor model selected |
| Logistic Regression | Discrimination | ROC / AUC | AUC > 0.90 |
| Logistic Regression | Multicollinearity | VIF via auxiliary lm | wt and hp: VIF < 2, no concern |
| Multinomial | Predictive accuracy | Confusion matrix | 96.9% |
| LDA | Normality assumption | Q-Q plots | Approximately Normal |
| LDA | Discrimination | ROC / AUC | AUC > 0.90 |
| All classifiers | Generalisation | CV (LOOCV / 10-fold) | Stable performance |

```{r diag-cooks, fig.cap="Cook's distance for the linear regression model. No observation exerts extreme leverage."}
cooksd <- cooks.distance(lm_step)
data.frame(obs = seq_along(cooksd), cd = cooksd,
           car = rownames(mtcars)) |>
  ggplot(aes(x = obs, y = cd)) +
  geom_segment(aes(xend = obs, yend = 0), colour = "#dee2e6") +
  geom_point(aes(colour = cd > 4 / nrow(mtcars)), size = 2.5) +
  geom_hline(yintercept = 4 / nrow(mtcars),
             linetype = "dashed", colour = col_red, linewidth = 0.8) +
  geom_text(aes(label = ifelse(cd > 4 / nrow(mtcars), car, "")),
            vjust = -0.6, size = 2.8, colour = col_red) +
  scale_colour_manual(values = c("FALSE" = col_blue, "TRUE" = col_red),
                      labels = c("Normal", "Influential"), name = NULL) +
  labs(title    = "Cook's Distance — Linear Regression",
       subtitle = paste0("Dashed line = 4/n threshold (",
                         round(4 / nrow(mtcars), 3), ")"),
       x = "Observation Index", y = "Cook's Distance") +
  theme_portfolio()
```

---

# Summary {#summary}

> **Objective 4:** Communicating findings.

## What I Did

I analysed performance data from 32 cars tested in 1973--74 using five
statistical methods to understand what drives fuel efficiency, how to classify
transmission type, and how to predict the number of cylinders.

## Key Findings

**What predicts fuel efficiency? (Linear Regression, LASSO, Ridge)**

Car weight is by far the strongest predictor, every additional 1,000 lbs
reduces fuel efficiency by about 3.9 mpg, holding other factors constant.
Horsepower also hurts efficiency, while a longer quarter-mile time (indicating
a less aggressive engine setup) is associated with better mpg. The model
explains `r round(summary(lm_step)$adj.r.squared * 100, 1)`% of the variation
in fuel consumption. LASSO and Ridge regularisation confirmed these findings
are robust and not an artefact of overfitting.

**Can we classify transmission type? (Logistic Regression, LDA)**

Both logistic regression and LDA correctly identified transmission type in over
`r round(max(log_acc, lda_acc) * 100, 0)`% of cars. Heavier cars are strongly
associated with automatic transmission, each 1,000 lb increase in weight
dramatically reduces the odds of having a manual gearbox. The AUC of
`r round(auc_log, 3)` confirms excellent discriminative ability.

**What predicts number of cylinders? (Multinomial Regression)**

Weight, horsepower, and displacement together predict cylinder count with
`r round(multi_acc * 100, 1)`% accuracy. Light, low-powered cars with small
engines are overwhelmingly likely to have 4 cylinders, while heavy,
high-powered cars almost exclusively have 8.


## Limitations

- Very small sample n = 32 observations, 1973--74 cars), results may not generalise to
  modern vehicles.
- Many predictors are highly correlated (weight, displacement, horsepower),
  making precise individual effect estimates difficult.

---

# Conclusion & Objectives Reflection {#conclusion}

> **All Objectives:** A reflective summary of how each learning objective was
> addressed through the analysis conducted in this portfolio.

## Objectives Achievement Summary

```{r conclusion-table,echo=FALSE}
tribble(
  ~`#`, ~Objective, ~`Method(s)`, ~`How Demonstrated`,
  1, "Probability & MLE foundations",
     "All models",
     "Normal, Bernoulli, and Multinomial distributions selected to match each outcome type. Log-likelihood visualised for the Normal model fitted to mpg, showing the MLE as the peak. AIC, BIC, and LRT all derived from the log-likelihood and applied throughout.",
  2, "Apply appropriate GLM",
     "Linear, Logistic, Multinomial, LDA",
     "OLS/Normal for continuous mpg; Binomial logit for binary am; Multinomial logit for three-class cyl; LDA as a generative alternative for am. Each choice justified by the distributional nature of the outcome.",
  3, "Model selection",
     "AIC, BIC, LRT, stepAIC(), LOOCV, 10-fold CV, LASSO lambda",
     "Candidate model comparison using AIC and BIC for linear regression; LRT confirming the core logistic model; stepwise selection applied after multicollinearity resolution; LOOCV for regression and 10-fold CV for classification; lambda tuned by cross-validation for LASSO and Ridge.",
  4, "Communicate to general audience",
     "All result sections",
     "Explanations accompany every model output. The separation plot and odds ratio forest chart communicate logistic regression results visually. The predicted probability curves show multinomial findings without equations. A dedicated plain language summary section translates all findings for a non-technical reader.",
  5, "R implementation",
     "lm, glm, glmnet, nnet, MASS, caret, pROC, car, lmtest, broom",
     "All models fitted, diagnosed, and compared entirely in R. Reproducible code throughout. tidy() used for clean coefficient output. ggplot2 for all visualisations with a consistent custom theme."
) |>
  kbl(row.names = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) |>
  column_spec(1, bold = TRUE, width = "2em") |>
  column_spec(2, bold = TRUE, width = "14em")
```

## Reflections on Each Objective

**Objective 1: Probability as a Foundation**

This objective was addressed from the very start of the modelling process. Before
fitting any model, I explicitly identified the probability distribution appropriate
for each outcome: Normal for `mpg`, Bernoulli for `am`, and Multinomial for `cyl`.
The log-likelihood curve for the Normal model (Section 3) demonstrated visually
that the MLE is the value of mean that maximises the log-likelihood, in this
case the sample mean. Throughout the portfolio, AIC, BIC, and LRT were used as
model comparison tools, all of which are derived directly from the
log-likelihood at the MLE. This established probability theory not as background
reading, but as the active framework underpinning every modelling decision.

**Objective 2: Applying the Appropriate GLM**

Three distinct outcome types required three distinct modelling approaches. Linear
regression was chosen for `mpg` because it is continuous and approximately
symmetric, the Normal error assumption is justified and OLS is the MLE. Logistic
regression was chosen for `am` because it is binary and the Bernoulli distribution
with logit link is the canonical GLM for two-class outcomes. Multinomial regression
was chosen for `cyl` because the three cylinder categories are unordered,
treating them as numeric would impose an arbitrary equal spacing. LDA was also
applied to `am` as a generative alternative, achieving a comparable AUC of
`r round(auc_lda, 3)` to the logistic model's `r round(auc_log, 3)`. In each
case the choice of model was driven by the statistical properties of the outcome,
not convenience.

**Objective 3: Model Selection**

Model selection was demonstrated through multiple complementary approaches. For
linear regression, a table of four candidate models was compared using AIC, BIC,
adjusted R-squared, and RMSE, the Full model achieved the lowest AIC and was
taken forward. Stepwise selection (`stepAIC`) was then applied and, critically, a
multicollinearity diagnosis revealed that `disp` had a VIF exceeding 10. Dropping
`disp` and re-running stepwise produced a final model with all VIF values below 5.
For logistic regression, the LRT confirmed that `wt + hp` significantly improved
on `wt` alone, while adding a third predictor caused complete separation, a
concrete illustration of why model selection must account for sample size
constraints. LOOCV was used to compare OLS, LASSO, and Ridge regression,
and 10-fold CV compared logistic regression against LDA. LASSO lambda was tuned
by cross-validation across 40 grid values. Across all methods, model selection
was principled and evidence-based.

**Objective 4: Communicate to a General Audience**

Every section contains an explanation block that translates model
output into interpretable statements. Coefficients are expressed as real-world
effects (e.g., "every additional 1,000 lbs reduces mpg by approximately 3.9").
Odds ratios are presented in a forest plot with a clear reference line at 1,
making direction and magnitude immediately readable. The separation plot shows
visually why weight alone almost perfectly distinguishes automatic from manual
cars. The predicted probability curve for multinomial regression communicates
the model's output without any equations. The dedicated Plain Language Summary
section (Section 9) consolidates all findings for a non-specialist reader,
and avoids statistical jargon throughout.

**Objective 5: R Implementation**

The entire workflow was implemented in R using a consistent and reproducible
coding style. The `broom` package was used throughout to produce tidy, tabular
model outputs via `tidy()` and `glance()`. `ggplot2` with a custom `theme_portfolio()`
function was used for all visualisations, ensuring a consistent visual style.
`glmnet` was used for LASSO and Ridge with cross-validation via `cv.glmnet()`.
`nnet::multinom()` was used for multinomial regression with manually computed
z-scores and p-values. `MASS::lda()` was used for discriminant analysis.
`caret::train()` with `trainControl()` was used for LOOCV and 10-fold CV.
Multicollinearity was diagnosed and resolved using `car::vif()`. Model
assumptions were tested with `lmtest::bptest()` and `ResourceSelection::hoslem.test()`.
ROC curves and AUC were computed with `pROC`. All code is visible in the
rendered document via code folding, making the analysis fully reproducible.

---

*Portfolio produced using R `r R.version$major`.`r R.version$minor` and RMarkdown.*
*Data: Henderson & Velleman (1981).*
